Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2016-07-01 docs/ms_sageAbundance.tex
adds tex verision of ms

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

%%% Change title format to be more compact
\usepackage{titling}
\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\section{Forecasting climate change impacts on plant population dynamics
at large spatial extents: a test case with sagebrush (\emph{Artemisia})
species}\label{forecasting-climate-change-impacts-on-plant-population-dynamics-at-large-spatial-extents-a-test-case-with-sagebrush-species}

\subsubsection[Andrew T. Tredennick, Mevin B. Hooten, Cameron L.
Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric
LaMalfa, Rebecca Mann, and Peter B. Adler]{Andrew T.
Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}},
Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles,
Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B.
Adler}\label{andrew-t.-tredennickcorrauth-mevin-b.-hooten-cameron-l.-aldridge-colin-g.-homer-david-t.-iles-andrew-kleinhesselink-eric-lamalfa-rebecca-mann-and-peter-b.-adler}

\emph{Andrew T. Tredennick, Department of Wildland Resources and the
Ecology Center, Utah State University, Logan, UT}

\emph{Mevin B. Hooten, U.S. Geological Survey, Colorado Cooperative Fish
and Wildlife Research Unit; Department of Fish, Wildlife, and
Conservation Biology, Colorado State University; Department of
Statistics, Colorado State University; Graduate Degree Program in
Ecology, Colorado State University, Fort Collins, CO}

\emph{Cameron L. Aldridge, Department of Ecosystem Science and
Sustainability, Colorado State University; Natural Resource Ecology
Laboratory, Colorado State University; U.S. Geological Survey, Fort
Collins Science Center; Graduate Degree Program in Ecology, Colorado
State University, Fort Collins, CO}

\emph{Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources
Observation and Science (EROS) Center, Sioux Falls, SD}

\emph{David T. Iles, Department of Wildland Resources and the Ecology
Center, Utah State University, Logan, UT}

\emph{Andrew Kleinhesselink, Department of Wildland Resources and the
Ecology Center, Utah State University, Logan, UT}

\emph{Eric LaMalfa, Department of Wildland Resources and the Ecology
Center, Utah State University, Logan, UT}

\emph{Rebecca Mann, Department of Wildland Resources and the Ecology
Center, Utah State University, Logan, UT}

\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
Center, Utah State University, Logan, UT}

\subsection{Summary}\label{summary}

\begin{enumerate}
\item Global environmental changes, like climate change, tend to play out at the landscape and regional scales. Thus, it is reasonable that attempts to forecast the impacts of such change do so at similar spatial scales. For plant species, this has meant relying primarily on a single tool, species distribution models, that contain little information on population dynamics and states.

\item Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using “local” scale data from, generally, 1 $\text{m}^{2}$ plots. These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.

\item Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model. At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals. To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (Artemesia spp.) cover in southwestern Wyoming. We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.

\item The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.

\item Our approach...
\end{enumerate}

\emph{Keywords}: population model, forecasting, integral projection
model, etc.

@@

Forecasting the impacts of climate change on plant populations and
communities is a central challenge of modern ecology. The challenge lies
in faithfully representing population dynamics at spatial and temporal
scales relevant to policy and management decisions. However, almost
every study of plant population dynamics relies on demographic
observations recorded at the meter to sub-meter scale. Local-scale
demographic data make building population projection models an easy task
(Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is
very difficult to extrapolate small-scale studies to large spatial
extents with any certainty because the data likely only represent a
small subset of parameter space and environmental conditions. Also,
given the costliness (in time and money) of collecting detailed
demographic data, it is unlikely that ``enough'' local-scale data can be
collected to accurately project large-scale population dynamics.

The fundamental limitation of local-scale data is that they are
insufficient for estimating the full extent of variation in model
parameters possible at larger spatial and temporal extents (Freckleton
et al., 2011; Queenborough et al., 2011). It is possible to estimate the
variance of population model parameters related to spatial and temporal
variability, but the observations used to estimate parameter uncertainty
tend to be clustered in space or time. Thus, extrapolating beyond the
spatial or temporal resolution of the data used to parameterize models
can lead to biased and wrong forecasts of future population states.

Alternatively, large-scale trends in populations are easily detected
using widely available monitoring data, but such data are rarely used to
project future population states. Given the tradeoffs between the
spatial-scale of data and the inference that can be drawn from the data,
we suggest combining the best features of local-scale population
modeling and large-scale monitoring data. Specifically, we propose a new
approach for modeling population dynamics at large spatial extents based
on the theory and mechanics of individual-based models (IBMs) as applied
to remotely-sensed time series data.

The approach we describe in this paper flows naturally from recent work
on density-structured population models. Density-structured models rely
on modeling the transitions of a population among discrete states rather
than the traditional approach of modeling the transitions of
individuals. Relatively few ecologists have taken up the
density-structured mantle first proposed by Taylor, C. M. et al. (2004),
but recently Freckleton et al. (2011) and Queenborough et al. (2011)
have higlighted the potential value the approach.

@@

\subsubsection{Data: Remotely-sensed time
series}\label{data-remotely-sensed-time-series}

To show our pixel-based modeling approach we use a subset of a
remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub
percent cover in Wyoming (Homer et al., 2012) (Figure
{[}fig:Location-of-the{]}). Sagebrush percent cover was estimated using
a regression tree to relate ground reflectances retrieved by three
sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780
field observations of sagebrush cover distributed across Wyoming. The
regression tree model was further validated using another 500 field
observations. For Wyoming sagebrush, the model achieved an \(R^{2}\) =
0.65. To hind-cast sagebrush cover the regression tree model was applied
to historical remote sensing images. This resulted in an annual time
series of sagebrush cover at 30 m resolution from 1984 to 2011. In this
dataset, values represent the percentage of a 30 \(\times\) 30 m pixel
covered by sagebrush. For our purposes we use two 5 \(\times\) 5 km
subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each
full dataset contains 754,596 observations (27 years \(\times\) 27,948
pixels).

\subsubsection{Data: Climate covariates}\label{data-climate-covariates}

Our approach is based on modeling how changes in plant populations
relate to climate variables. We a priori narrowed our focus to climate
covariates we know are important for sagebrush and that could be easily
calculated from general circulation model future projections. In all
vital rate regressions we focus on the demographic processes driving
changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t},
and how these processes are mediated by climate. The four climate
variables in our vital rate regressions are: (1) year \emph{t} fall
through summer precipitation, (2) year \emph{t}-1 fall through summer
precipitation, (3) year \emph{t} average spring temperature, and (4)
year \emph{t}-1 average spring temperature (Table
{[}tab:Definition-of-climate{]}). Climate data to parameterize vital
rate regressions (`observed' climate) came from the PRISM climate
database. Climate data were extracted for the centroid our study area.

Climate data used to inform model projections of future sagebrush cover
came from average outputs of all models used in the most recent IPCC
report, the Coupled Model Intercomparison Project 5 (CMIP5;
\url{http://cmip-pcmdi.llnl.gov/cmip5/}). We downloaded model estimates
of historic climate (1950-2000) and projected climate (2050-2100) for
the centroid of our study area. We wanted to keep the temporal
variability from observed climate consistent, so our aim was to adjust
observed climate according to projected changes. For temperature we
simply subtracted the average historic temperature from the average
projected temperature. For precipitation we calculated the proportional
change: we subtracted average projected precipitation from average
historic precipitation, and then divided by average historic
precipitation. We did this for three ``Representative Concentration
Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
\url{http://tntcat.iiasa.ac.at/RcpDb/}. Basically they correspond to
stabilization of radiative forcing before 2100, after 2100, and
increasing greenhouse gas emissions, respectively. We implemented
projected temperature and precipitation changes in the model by
perturbing observed annual values by the values in Table \#.

\begin{table}[h]
\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
average projections.}
\begin{tabular}{llllll}
\hline
 & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
\hline
RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
\hline
\end{tabular}
\end{table}

@@

We use a descriptive model for sagebrush cover that includes additive
spatial and temporal effects. Since the spatial and temporal effects are
additive, we are not considering space-time interactions. Our model aims
to relate observed sagebrush cover, \(C_{s,t}\), at time \emph{t} and
spatial location \emph{s} to an underlying process \(\mu_{s,t}\) that is
defined over all spatial locations and observation times. For
statistical convenience, we consider discrete values of sagebrush
percent cover as counts, giving us the formulation

\begin{align}
[\textbf{C}] &= \text{Poisson}(\boldsymbol{\lambda}) \\
\boldsymbol{\lambda} &= \text{exp}(\boldsymbol{\mu}) 
\end{align}

which states that counts (integer percent cover) are Poisson distributed
with a log link function on abundance intensity
(\(\boldsymbol{\lambda}\)). We assume the underlying process,
\(\boldsymbol{\mu}\), is a function of climate covariates (\textbf{X}),
lagged sagebrush abundace (a temporal process), and a purely spatial
random effect (\(\boldsymbol{\eta}\)):

\begin{align}
\mu_{s,t} = \underbrace{\beta_{0} + \beta_{1}\mu_{s,t-1}}_\text{temporal} + \underbrace{\mathbf{X}_{t}\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{s}}_\text{spatial} + \underbrace{\varepsilon_{s,t}}_\text{error}
\end{align}

Fitting the spatial random effect (\(\eta\)) is computationally
demanding when the observations come from 10s of 1,000s locations. The
dimensionality is just too large. To overcome these computational
constraints we employed a dimension reduction strategy to reduce the
number of parameters that need to be estimated to account for spatial
variation. A common strategy is to express high dimensional effects,
like \(\boldsymbol{\eta}\), as the product of an expansion matrix,
\textbf{K}, and a smaller parameter vector, \(\boldsymbol{\alpha}\). We
can then define the spatial effect as

\begin{align}
\boldsymbol{\eta} = \textbf{K}\boldsymbol{\alpha}.
\end{align}

In this case, \(\boldsymbol{\alpha}\) is a \(m \times 1\) vector of
reduced spatial random effects, and \textbf{K} is a \(S \times m\)
matrix that maps the reduced effects to the full \emph{S}-dimensional
space, where \emph{S} is the total number of observed locations. Thus,
we are able to reduce the effective number of parameters from
\({S} \rightarrow m\).

Obviously, the last remaining obstacle is to parameterize \textbf{K}.
Here we use kernel convolution to interpolate the spatial random effect
between \emph{m} ``knots'' that are nonrandomly distributed across the
space of our study area. These means we are modeling spatial random
effects at the knot level, and we use \textbf{K} to interpolate those
effects between knots. We use a simple Gaussian kernel density to define
the distance-decay function around the knots (\textbf{w}), such that the
entries of \textbf{K} are

\begin{align}
K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
\end{align}

where \(d_{s,m}\) is the Euclidean distance between the centroid of
sample cell \emph{s} and the location of knot \emph{m}, and \(\sigma\)
is the kernel bandwidth. Based on residual spatial structure we found
after fitting a simple model without spatial effects, we set \(\phi\) to
about one-third of the range of spatial dependence evident in the
residual variograms -- 250 meters in this case. In the end, the simplest
way to understand our dimension reduction strategy is to consider
\(\alpha\)s as weights for each knot.

\subsubsection*{Computing and model
estimation}\label{computing-and-model-estimation}
\addcontentsline{toc}{subsubsection}{Computing and model estimation}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing
\usepackage[document]{ragged2e}
\usepackage{color, soul}
\usepackage{graphicx}

@@

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\begin{centering}
\textbf{\large{Forecasting climate change impacts on plant populations over large spatial extents}}

\textsc{\small{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B. Hooten\textsuperscript{2,3,4}, Cameron L. Aldridge\textsuperscript{5}, Collin G. Homer\textsuperscript{6}, Andrew R. Kleinhesselink\textsuperscript{1}, and Peter B. Adler\textsuperscript{1}}}

\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\
\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\
\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\
\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}} \\
\textit{\small{\textsuperscript{5}Natural Resource Ecology Laboratory and Department of Ecosystem Science and Sustainability, Colorado State University, Fort Collins, CO, in cooperation with US Geological Survey, Fort Collins Science Center, Fort Collins, CO}}\\
\textit{\small{\textsuperscript{6}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD 57198 USA}}

\end{centering}

\renewcommand*{\thefootnote}{\arabic{footnote}}

\setcounter{footnote}{0}

\subsection{Abstract}\label{abstract}

Plant population models are powerful tools for predicting climate change
impacts in one location, but are difficult to apply at landscape scales.
We overcome this limitation by taking advantage of two recent advances:
remotely-sensed, species-specific estimates of plant cover and
statistical models developed for spatio-temporal dynamics of animal
populations. Using computationally efficient model reparameterizations,
we fit a spatiotemporal population model to a 28 year time series of
sagebrush (\emph{Artemisia} spp.) percent cover over a \(2.5\times5\) km
landscape in southwestern Wyoming while formally accounting for spatial
autocorrelation. We include interannual variation in precipitation and
temperature as covariates in the model to investigate how climate
affects the cover of sagebrush. We then use the model to forecast the
future abundance of sagebrush at the landscape scale under projected
climate change, generating spatially explicit estimates of
\textcolor{blue}{sagebrush population trajectories} that have, until
now, been impossible to produce at this scale. Our broad-scale and
long-term predictions are rooted in small-scale and short-term
population dynamics and provide an alternative to predictions offered by
species distribution models that do not include population dynamics. Our
approach,
\textcolor{blue}{which combines several existing techniques in a novel way, demonstrates the use of}
remote sensing data to model population responses to environmental
change that play out at spatial scales far greater than the traditional
field study plot.

\emph{Key words: population model, climate change, forecasting,
spatiotemporal model, remote sensing, sagebrush}, Artemisia,
\emph{dimension reduction}

@@

Forecasting the impacts of climate change on plant populations and
communities is a central challenge for ecology (Clark et al. 2001,
Petchey et al. 2015). Population models are ideally suited for meeting
such a challenge because they provide a way to link climate drivers
directly to population dynamics (Hare et al. 2010, Adler et al. 2012,
Ross et al. 2015, Shriver 2015). However, inference from population
models is typically limited to small spatial extents because the data
required is difficult to collect across broad species ranges. Almost
every study of plant population dynamics relies on demographic
observations recorded at the meter to sub-meter scale (see, e.g.,
Salguero-G{ó}mez et al. 2015). Local-scale demographic data make
building population projection models an easy task (Ellner and Rees
2006, Rees and Ellner 2009, Adler et al. 2012), but it is very difficult
to extrapolate small-scale studies to large spatial extents with any
certainty because the data likely only represent a small subset of
parameter space and environmental conditions (Freckleton et al. 2011,
Queenborough et al. 2011). The real challenge is not to simply make
population forecasts, but to do so at spatial scales relevant to policy
and management decisions (Queenborough et al. 2011).

The ideal tool would be a \textcolor{blue}{broad}-scale, dynamic
population model (Schurr et al. 2012, Merow et al. 2014), but developing
useful models at this scale has been limited by the availability of time
series data at large spatial extents and statistical methods for fitting
high-dimensional spatial models. Fortunately, new advances in remote
sensing and statistics now allow us to overcome both of these
limitations. First, new remote-sensing (RS) methods are now producing
accurate time series of species-specific plant cover at landscape
scales. These data can be fit with dynamic population models which
include yearly fluctuations in climate as covariates. Such RS time
series have revolutionized models of how climate affects ecosystem-level
processes (e.g., Running et al. 2004) and have been used to detect
long-term trends in plant population abundance (e.g., Homer et al.
2015), but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed dimension reduction
and reparameterization techniques to efficiently fit high-dimension
spatiotemporal models (see Conn et al. 2015 for a review). These new
statistical methods have yet to be applied to RS-derived plant
population data at broad scales.

Large-scale, spatially-explicit population models based on RS data could
offer a valuable new way to investigate the effects of large-scale
environmental changes playing out at landscape and regional scales. Most
current assesments of how plant and animal populations will respond to
climate change rely on species distribution models (SDMs). SDMs rely on
static associations between contemporary climate and a species'
distribution or, more rarely, abundance to project future distribution
or abundance (Elith and Leathwick 2009) and they are easily applied at
landscape to continental scales (e.g., Maiorano et al. 2013, Clark et
al. 2014). However, the short-term and small-scale population dynamics
that actually drive the large-scale distributions of species are not
represented in most SDMs. Because SDMs typically rely on occurence data,
their projections of habitat suitability or probability of occurence
provide little information on the future states of populations in the
core of their range -- areas where a species exists now and is expected
to persist in the future (Ehrl{é}n and Morris 2015). Furthermore,
because they lack short-term dynamics, SDMs usually cannot produce any
estimate of the rate at which local populations will increase or
decrease in the near-term and instead project a future equilibrium
species distribution that may or may not ever be reached. Direct
validation of such predictions is extremely rare (Roberts and Hamann
2012). Large-scale dynamic population models could overcome these
limitations. They would produce spatially-explicit estimates of species
abundance within the species range (Ehrl{é}n and Morris 2015), have the
potential to model expansion in abundance outside the range when coupled
with dynamic models of dispersal, and would provide testable predictions
of how populations should respond to short-term climate perturbations.
These short-term predictions also would give modelers the opportunity to
repeatedly validate and refine their models (Luo et al. 2011).

Sagebrush (\emph{Artemisia} spp.) ecosystems offer an ideal testing
ground for new spatially explicit population models derived from RS
data. Sagebrush species are widely distributed (Kuchler 1964), they are
sensitive to climate (Perfors et al. 2003, Miglia et al. 2005, Poore et
al. 2009, Dalgleish et al. 2011, Xian et al. 2012, Apodaca 2013,
Schlaepfer et al. 2014a, 2014b, Harte et al. 2015, Homer et al. 2015),
new landscape and regional scale time series of sagebrush cover are now
being produced from aerial imagery (Homer et al. 2012), and forecasts of
future sagebrush ecosystems are in high demand due to the precarious
conservation status of the greater sage-grouse (\emph{Centrocercus
urophasianus}) (Arnett and Riley 2015). SDMs typically predict that much
of the area occupied by sagebrush ecosystems today will become
unsuitable for sagebrush due to climate change, resulting in a dramatic
loss in the extent of sagebrush habitat by the end of this century
(Shafer et al. 2001, Neilson et al. 2005, Bradley 2010, Schlaepfer et
al. 2012, Still and Richardson 2015). Ecohydrology models supply a
possible mechanism for sagebrush losses predicted by SDMs: climate
warming could lead to earlier snowmelt, increased evaporation and
ultimately less recharge of deeper soil layers in the spring (Schlaepfer
et al. 2012, 2014a). In warmer parts of its range, increased temperature
could be especially detrimental to sagebrush as it depends on water from
deeper soil to survive and grow in this arid region (Pechanec et al.
1937, Schlaepfer et al. 2011, Germino and Reinhardt 2014). In contrast,
at higher elevations and in colder regions, warming and earlier snowmelt
could lengthen the growing season and increase sagebrush occurrence
(Schlaepfer et al. 2012, 2014a). Direct observations of individual
plants and experimental plots tend to agree with these models: growth
tends to respond negatively to spring and summer temperatures (Miglia et
al. 2005, Poore et al. 2009, Apodaca 2013) except at higher elevations
where earlier snowmelt may allow for a longer growing season (Perfors et
al. 2003, Harte et al. 2015). A large-scale, spatially-explicit
population model for sagebrush driven by interannual climate variability
would provide a valuable new tool for assessing how sagebrush could
respond to climate change in the future.

Building on recent technological advances in spatial statistics (Latimer
et al. 2009, Conn et al. 2015) and anticipating ever-increasing
availability of RS data (He et al. 2015), we demonstrate how large-scale
plant population models could be used to predict population impacts of
climate change. \textcolor{blue}{As a proof-of-concept, we use} a
process model motivated by Gompertz density-dependent population growth
and a remotely-sensed time series of sagebrush cover from Wyoming (Homer
et al. 2012, 2015). We account for spatial autocorrelation with
dimension reduction techniques (Latimer et al. 2009, Conn et al. 2015)
and produce spatially-explicit estimates of sagebrush percent cover.
\textcolor{blue}{Unlike most SDMs, our approach models the dynamics of plant abundance through time, and thus, is a population model, in the same spirit that models of animal counts through time are population models.}
The modeling framework we propose can be applied to any
spatially-explicit time series of plant cover or density, but its
application to remotely-sensed data products offers the greatest
potential to combine the information of population models (e.g.,
population status \textcolor{blue}{and temporal dynamics}) and the
spatial extent of species distribution models.

@@

\subsubsection{Data}\label{data}

\paragraph{Remotely-sensed time
series}\label{remotely-sensed-time-series}

To demonstrate our modeling approach, we use a subset of a
remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy
cover in Wyoming (Homer et al. 2012). As part of a separate study,
\textcolor{blue}{Homer et al. 2012 estimated} sagebrush percent cover
using a regression tree to relate ground reflectances retrieved by three
sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780
field observations of sagebrush cover distributed across Wyoming. The
regression tree model was further validated using another 297 field
observations. For Wyoming sagebrush, the model achieved an \(R^{2}\) =
0.65 and an out-of-sample RMSE of 5.46\% (Homer et al. 2012). To
hind-cast sagebrush cover the regression tree model was applied to
historical remote sensing images
\textcolor{blue}{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
This resulted in an annual time series of sagebrush cover at 30 meter
resolution from 1984 to 2011 (Fig. B1). In this
\textcolor{blue}{remote sensing product}, values represent the
percentage of a 30 \(\times\) 30 meter pixel covered by sagebrush. In
our study, we focused on a 5,070 \(\times\) 2,430 meter subset totaling
13,689 30 \(\times\) 30 meter pixels each year (Fig. 1). Thus, the
subset of the \textcolor{blue}{remote sensing product} we use contains
369,603 observations spanning 27 year-to-year transitions (27 years
\(\times\) 13,689 pixels).

\paragraph{Climate covariates}\label{climate-covariates}

Our approach models interannual changes in plant cover as a function of
seasonal climate variables. We used daily historic weather data for the
center of our study site from the NASA Daymet data set (\emph{available
online})\footnote{\url{http://daymet.ornl.gov/}}. The Daymet weather
data are interpolated between coarse observation units and capture some
spatial variation. We relied on weather data for the centroid of our
study area. We calculated five climate variables from the Daymet data
for the time period coinciding with our remotely sensed data (1984 to
2011).

We narrowed our focus to climate covariates we know are important for
sagebrush and that could be calculated from general circulation model
projections. The five climate variables in our population model are: (1)
cumulative, ``water year'' precipitation for year \emph{t}-2
(\emph{lagPpt}), (2) year \emph{t}-1 fall through summer precipitation
(\emph{ppt1}), (3) year \emph{t} fall through summer precipitation
(\emph{ppt2}), (4) year \emph{t}-1 average spring temperature
(\emph{TmeanSpr1}), and (5) year \emph{t} average spring temperature
(\emph{TmeanSpr2}), where \emph{t-1} to \emph{t} is the transition of
interest. We selected these variables \emph{a priori} based on previous
studies (see \emph{Introduction}), though not all emerge as important
predictors in our model.

@@

We use a descriptive model for sagebrush cover that includes additive
spatial and temporal effects similar to that described by Conn et al.
(2015). Interannual change in percent cover represents the integrated
outcome of recruitment, survival, growth, and retrogression (shrinkage)
of individual plants from year to year. We model observed integer
percent cover (\emph{y}) in cell \emph{i} at time \emph{t} as
conditionally Poisson

\vspace{-4em}\begin{align}
y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
\end{align}\vspace{-4em}

where \(\mu_{i,t}\) is the expected percent cover of pixel \emph{i} in
year \emph{t}

\vspace{-4em}\begin{align}
\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal + dens. dep} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
\end{align}\vspace{-3em}

Our model of percent cover change includes a density-dependent effect of
log-transformed cover in the previous year (\(y_{i,t-1}\)), climate
effects (\(\textbf{x}_t\)), and a spatial random effect (\(\eta\)) for
each pixel \emph{i}. Climate effects were standardized
{[}\((x_{i}-\bar{x}) / \sigma(x)\){]} to improve convergence during the
model fitting stage and to allow for easier prior specification. The
intercept, \(\beta_{0,t}\), was allowed to vary through time; these
random year effects recognize that all observations from a particular
year share the same climate covariates and thus are not independent. We
used a Poisson likelihood because integer percent cover values in the
sagebrush data product can be considered a form of count data. We also
evaluated a negative binomial model, but found little evidence for
overdispersion beyond what our model was already accomodating via the
spatial random effects (\(\boldsymbol{\eta}\)). There was no evidence of
zero-inflation in our data, but see below (\emph{Accomodating zeros})
for how we handled the small number of zero percent cover observations.
\textcolor{blue}{We assume that the remotely sensed estimates of percent cover are "true" and free of error.}
\textcolor{blue}{This need not be the case, and if measurement or sampling error is known then it could be included in our Bayesian model as a "sampling model"}
(Hobbs and Hooten 2015).

The spatial random effect (\(\boldsymbol{\eta}\)) accounts for spatial
autocorrelation among pixels that occur near each other in space. Thus,
\(\boldsymbol{\eta}\) acts as an offset on the intercept
(\(\beta_{0,t}\)), creating a spatial field that defines how pixels
differ from the mean, on average, in space (e.g., areas of perennially
low or high cover, relative to average cover). Fitting the model with a
spatial random effect (\(\eta\)) is computationally demanding for large
data sets like ours. The computational demand is due to the required
calculations of the spatial covariance matrices, which increase as a
cubic function of the number of locations (Wikle 2010). Key to our
approach is a dimension reduction strategy that greatly reduces the
number of parameters needed to be estimated to account for spatial
variation by reducing the size of the spatial covariance matrices that
need to be inverted at each MCMC iteration. Fitting models that
appropriately account for spatial autocorrelation over large spatial
extents would not be feasible without these modern techniques. Our
dimension reduction strategy expresses the high dimensional spatial
random effect, \(\boldsymbol{\eta}\), as the product of an expansion
matrix, \textbf{K}, and a smaller parameter vector,
\(\boldsymbol{\alpha}\) (e.g., Hooten et al. 2003, Hooten and Wikle
2007, Conn et al. 2015). We can then approximate the spatial effect as

\vspace{-4em}\begin{align}
\boldsymbol{\eta} &\approx \textbf{K}\boldsymbol{\alpha}, \\
\alpha_{m} &\sim \text{Normal}(0,\sigma_{\eta}^2).
\end{align}\vspace{-4em}

In this case, \(\boldsymbol{\alpha}\) is a \(m \times 1\) vector of
reduced spatial random effects, and \textbf{K} is a \(S \times m\)
matrix that maps the reduced effects to the full \emph{S}-dimensional
space, where \emph{S} is the total number of observed locations. Thus,
we are able to reduce the effective number of parameters from \({S}\) to
\(m\).

The last remaining obstacle is to parameterize the matrix of basis
functions, \textbf{K}. We use kernel convolution (Barry and Hoef 1996,
Higdon 1998) to interpolate the spatial random effect between \emph{m}
``knots'' that are nonrandomly distributed across the space of our study
area. This means we are modeling spatial random effects at the knot
level, and we use \textbf{K} to interpolate those effects between knots.
We use an exponential kernel density to define the distance-decay
function around the knots (\textbf{w}), such that the entries of
\textbf{K} are

\vspace{-4em}\begin{align}
K_{s,m} &= w_{s,m} / \sum_{s=1}^{S} w_{s,m}
\end{align}\vspace{-4em}

where

\vspace{-4em}\begin{align}
w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right)
\end{align}\vspace{-4em}
